Metabolites reflect variability introduced by mesoscale eddies in the North Pacific Subtropical Gyre

### 2023-11-11 manuscript/manuscript.Rmd
Finished adding notes and comments from Anitra

Despite its historical reputation as a static and unchanging environment, the North Pacific Subtropical Gyre shows significant variation in its particulate composition due to mesoscale eddies. We sampled the metabolome across two sets of eddy dipoles to explore the total effect of sea level anomaly (SLA).

In all of the datasets analyzed here, we detected a significant difference in the composition of the metabolome. This effect was strongest in the samples taken during the Lagrangian stations at the center of each eddy in the 2017 MESOSCOPE cruise, with nearly a quarter of the total variance explained by the eddy from which the samples were taken. In the samples taken along the eddy transect, the largest effect was detected in the deepest samples taken from 175 meters and from the DCM, with much less of a response at the surface. In the Falkor samples, however, the opposite was true with a larger SLA response at the surface than that at the DCM. In each case, the differences introduced by SLA were smaller than those of depth when compared directly, with even the high-resolution sampling around the DCM finding slightly more variance explained by the 40 meter difference in sampling depth than the 40 centimeter difference in SLA. This result agrees well with prior research by @Heal2021 and @Kumler2023 showing that depth has a large effect on the metabolome.

@@

```{r multimodel testing maybe, eval=FALSE}

dist_mat <- untarg_peaks %>%
  left_join(filled_file_metadata %>% distinct(ms_run, station, filename, depth), 
            by="filename") %>% filter(ms_run=="MC") %>% 
  mutate(depth=factor(depth, levels=c("DCMless20m", "DCMless10m", "DCM", 
                                      "DCMplus10m", "DCMplus20m"))) %>%
  arrange(feature, station, depth) %>%
  mutate(filename=fct_inorder(filename)) %>%
  group_by(feature) %>%
  mutate(norm_area=rank(norm_area)) %>%
  select(feature, filename, norm_area) %>%
  pivot_wider(names_from=feature, values_from="norm_area") %>%
  mutate(filename=str_remove_all(filename, "180205_Smp_|.mzML")) %>%
  column_to_rownames("filename") %>%
  dist(method = "manhattan") %>%
  as.matrix()
dist_mat %>%
  as.data.frame() %>%
  rownames_to_column("filename_a") %>%
  pivot_longer(-filename_a, names_to="filename_b", values_to = "dist") %>%
  mutate(filename_a=fct_inorder(filename_a)) %>%
  mutate(filename_b=fct_inorder(filename_b)) %>%
  filter(as.numeric(filename_b)>=as.numeric(filename_a)) %>%
  ggplot() +
  geom_raster(aes(x=filename_a, y=filename_b, fill=dist)) +
  coord_fixed() +
  scale_fill_gradient2(midpoint = 2200) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle=90, hjust=1, vjust=0.5))

@@

The NMDS plots of the high-resolution depth samples were also illustrative of sample similarity, with clear SLA and depth trends. The anticyclonic samples grouped tightly and showed little variance with depth or triplicate when compared to the cyclonic samples with the exception of DCM triplicate B, which was highly distinct for unknown reasons. The cyclonic samples, on the other hand, were much more variable. One notable aspect of their clustering was the way in which the deepest samples (DCM plus 20m) grouped most closely to the anticyclonic samples, perhaps indicating that below the DCM the metabolome rapidly approaches a uniform deep-water signal as seen in the 175 meter samples of the eddy transect in @Kumler2023, which showed that the samples from the deep euphotic zone were more similar to each other than those taken from the surface or DCM were to other surface or DCM samples.

@@

The Falkor cruise upset many of the expectations we developed during the MESOSCOPE cruise in the previous year. Most surprising was the large difference between the eddies detected at the *surface*, while the DCM samples were functionally indistinguishable. This contrasts directly with the results from MESOSCOPE and the results in @Dugenne2023 and @Gleish2023, in which the largest differences were found at the DCM or below. This was especially surprising given the large inter-replicate differences between the Falkor surface samples, while the DCM samples tended to be much more visually consistent.

@@

Several molecule- or pathway-specific narratives emerged from the metabolomic data described above as well. The untargeted approach used here allowed us to describe and chatacterize signals from small molecules whose identity was unknown, a particularly promising approach in open ocean gyres where the largest fraction of unknowns is found [@Heal2021]. We found that this was especially true in anticyclonic eddies at all depths, with the diversity of the cyclone much better characterized by our list of authentic standards. Many of the trends detected below persisted even after normalizing within each sample, indicating that the shifts discussed have implications beyond simple scaling with biomass.

Of those molecules whose identity was known, the clearest response to SLA was that of the enigmatic osmolyte homarine at the DCM. This abundant compound increased approximately threefold in concentration from < 50 pM in the anticyclone to around 150 pM in the center of the cyclone in both the eddy transect and eddy center datasets. The pattern was weaker in the noisier Falkor data but the largest concentrations were still detected in the cyclone and the median cyclone measurement was greater than the maximum value for the anticyclone. This molecule has a well-established role as an osmolyte in eukaryotic phytoplankton and cyanobacteria [@Gebser2013; @Dawson2020; @Heal2021; @Durham2022] as well as a documented decrease in abundance with depth [@Heal2021], indicating that this shift is potentially due to any combination of differences in salinity, DCM depth, total biomass, and community composition. Homarine's tight correlation with trigonelline, however, is entirely mystifying as the two molecules are not well-known to have a strong biological relationship beyond their structural similarity.

@@

Similarly to homarine, isethionate is a known osmolyte found exclusively in eukaryotes according to a previous study [@Durham2022] and strongly associated with a few diatoms in particular [@ref or culture data]. This compound was enriched in the cyclone of the DCM in both the MESOSCOPE and Falkor cruises along with its precursor taurine, validating earlier results from @Barone2022 that showed strong enrichment of most eukaryotes at the DCM and *Pseudo-nitzschia* in particular. [Expand on this with after additional meetings with Bryn and following up with Sam - TPA (K03851 should definitely have been picked up, poly-A selected for euks?, see if iseJ (the 1.1.2-) can still be searched for without a KO number (yes according to Frank bc the sequence exists but complicated by catabolism/anabolism))]

@@

Taurine was also among the compounds highly enriched in the anomalous surface sample from Station 11 (MS11C215m_A), likely along with its trimethylated form taurine betaine. While putative in our analysis, trimethyltaurine has been previously reported in corals and red algae, where it probably has a destabilizing influence on proteins [@Impellizzeri1975; @Hill2017; @Bruzdziak2018]. We have a few hypotheses for the striking pattern in these compounds and the other 13 metabolites detailed above in Results. Most simply, the signal could simply be contamination during the sampling process, with some unknown material dominated by these compounds added to the filter during collection or processing. More interestingly, the signal could represent a real biological difference, perhaps indicating a pathway or organism highly abundant in this sample alone. As noted above, the MS11C215m_A sample was collected during a *Trichodesmium* bloom, an environment characterized by large heterogeneity even between replicate samples [@Rouco2014; @Rouco2016]. 10 liters of seawater were filtered for each sample but it is conceivable that a tuft or puff of *Trichodesmium* was captured in only this one as previously noted by @Rouco2014. The presence of a predictable trend across multiple chemically similar metabolites is the strongest bit of evidence in favor of the *Trichodesium* hypothesis, but the fact that these metabolites do not appear robustly in the bloom samples and that homoserine betaine did not appear to be especially abundant challenges it. One final note is that homoserine betaine was only documented as the sole major compatible solute in the Clade III *T. erythraeum* strain IMS101, while the bloom was more likely due to Clade I species given their relative abundance in the NPSG [@Rouco2016]. [Email Sonya about the presence of compatible solute genes in Clade I tricho? Especially taurine/TMT? Ask about IFCB samples (they're on the dashboard http://ifcbdb.soest.hawaii.edu/timeline?dataset=MESOSCOPE&bin=D20170708T043509_IFCB000 I just don't know what to ask)]

@@

A final pattern worth noting is the apparent upregulation of metabolites in the urea-ornithine cycle (OUC) among the samples taken during the Falkor cruise. Aspartate, citrulline, ornithine, and malic acid were all found to be more abundant in the Falkor samples relative to MESOSCOPE. The OUC is a staple of metazoan metabolism and was recently revealed to be present in stramenopiles as well, where it enables rapid response to increases in nitrogen supply [@Allen2011]. The enrichment of these metabolites in the Falkor cruise may indicate that there was still rapid uptake and turnover of nitrogen in the 2018 eddy pair while the 2017 eddies were more stable, despite both eddies being in a similar state of stable anticyclone and weakening cyclone [@Dugenne2023].

>>>>

The North Pacific Subtropical Gyre shows significant variation in its particulate composition due to mesoscale eddies. We sampled the metabolome across two sets of eddy dipoles to explore the effect of sea level anomaly (SLA).

In all of the datasets analyzed here, we detected a significant difference in the composition of the metabolome between eddies. This effect was strongest in the samples taken during the Lagrangian stations at the center of each eddy in the 2017 MESOSCOPE cruise, with nearly a quarter of the total variance explained by the eddy from which the samples were taken. In the samples taken along the eddy transect, the largest effect was detected in the deepest samples taken from 175 meters and from the DCM, with much less of a response at the surface. In the Falkor samples, however, the opposite was true with a larger SLA response at the surface than that at the DCM. In each case, the differences introduced by SLA were smaller than those of depth when compared directly, with even the high-resolution sampling around the DCM finding slightly more variance explained by the 40 meter difference in sampling depth than the 40 centimeter difference in SLA. This result agrees well with prior research by @Heal2021 and @Kumler2023 showing that depth has a large effect on the metabolome.

@@

```{r multimodel testing maybe, eval=FALSE}
dist_mat <- untarg_peaks %>%
  left_join(filled_file_metadata %>% distinct(ms_run, station, filename, depth), 
            by="filename") %>% filter(ms_run=="MC") %>% 
  mutate(depth=factor(depth, levels=c("DCMless20m", "DCMless10m", "DCM", 
                                      "DCMplus10m", "DCMplus20m"))) %>%
  arrange(feature, station, depth) %>%
  mutate(filename=fct_inorder(filename)) %>%
  group_by(feature) %>%
  mutate(norm_area=rank(norm_area)) %>%
  select(feature, filename, norm_area) %>%
  pivot_wider(names_from=feature, values_from="norm_area") %>%
  mutate(filename=str_remove_all(filename, "180205_Smp_|.mzML")) %>%
  column_to_rownames("filename") %>%
  dist(method = "manhattan") %>%
  as.matrix()
dist_mat %>%
  as.data.frame() %>%
  rownames_to_column("filename_a") %>%
  pivot_longer(-filename_a, names_to="filename_b", values_to = "dist") %>%
  mutate(filename_a=fct_inorder(filename_a)) %>%
  mutate(filename_b=fct_inorder(filename_b)) %>%
  filter(as.numeric(filename_b)>=as.numeric(filename_a)) %>%
  ggplot() +
  geom_raster(aes(x=filename_a, y=filename_b, fill=dist)) +
  coord_fixed() +
  scale_fill_gradient2(midpoint = 2200) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle=90, hjust=1, vjust=0.5))

@@

The NMDS plots of the high-resolution depth samples were also illustrative of sample similarity, with clear SLA and depth trends. The anticyclonic samples grouped tightly and showed little variance with depth or triplicate when compared to the cyclonic samples with the exception of DCM triplicate B, which was highly distinct for unknown reasons. In contrast, the cyclonic samples were much more variable as is perhaps expected when the biomass is in the form of larger phytoplankton that are less homogenous in the environment. One notable aspect of their clustering was the way in which the deepest samples (DCM plus 20m) grouped most closely to the anticyclonic samples, perhaps indicating that below the DCM the metabolome rapidly approaches a uniform deep-water signal as seen in the 175 meter samples of the eddy transect in @Kumler2023, which showed that the samples from the deep euphotic zone were more similar to each other than those taken from the surface or DCM were to other surface or DCM samples.

@@

The Falkor cruise upset many of the expectations we developed during the MESOSCOPE cruise in the previous year. Most surprising was the large difference between the eddies detected at the *surface*, while the DCM samples were functionally indistinguishable. This contrasts directly with the results from MESOSCOPE and the results in @Dugenne2023 and @Gleish2023, in which the largest differences were found at the DCM or below. This was especially surprising given the large inter-replicate differences between the Falkor surface samples, while the DCM samples tended to be much more consistent.

@@

Several molecule- or pathway-specific narratives emerged from the metabolomic data described above as well. The untargeted approach used here allowed us to describe and chatacterize signals from small molecules whose identity was unknown, a particularly promising approach in open ocean gyres where the largest fraction of unknowns is found [@Heal2021]. [Anitra: how many of your unknowns were also unknowns in Katherine's paper?] We found that this was especially true in anticyclonic eddies at all depths, with the diversity of the cyclone much better characterized by our list of authentic standards. Many of the trends detected below persisted even after normalizing within each sample, indicating that the shifts discussed have implications beyond simple scaling with biomass.

Of those molecules whose identity was known, the clearest response to SLA was that of the enigmatic osmolyte homarine at the DCM. This abundant compound increased approximately threefold in concentration from < 50 pM in the anticyclone to around 150 pM in the center of the cyclone in both the eddy transect and eddy center datasets. The pattern was weaker in the noisier Falkor data but the largest concentrations were still detected in the cyclone and the median cyclone measurement was greater than the maximum value for the anticyclone. This molecule has a well-established role as an osmolyte in eukaryotic phytoplankton and cyanobacteria [@Gebser2013; @Dawson2020; @Heal2021; @Durham2022] as well as a documented decrease in abundance with depth [@Heal2021], indicating that this shift is potentially due to any combination of differences in salinity [Anitra: Really? Did salinity change much? Will: Half a PSU at 175m, 0.3 at DCM. Idk if that's enough - check Hannah's paper?], DCM depth, total biomass, and community composition. Homarine was tightly correlated with trigonelline, and although they have structural similarity they are not known to have a relationship beyond their shared function as osmolytes.

@@

Similarly to homarine, isethionate is a known osmolyte thus far found exclusively in eukaryotes [@Durham2022] and strongly associated with a few diatoms in particular [@ref or culture data]. This compound was enriched in the cyclone of the DCM in both the MESOSCOPE and Falkor cruises along with its precursor taurine, validating earlier results from @Barone2022 that showed strong enrichment of most eukaryotes at the DCM and *Pseudo-nitzschia* in particular. [Expand on this with after additional meetings with Bryn and following up with Sam - TPA (K03851 should definitely have been picked up, poly-A selected for euks?, see if iseJ (the 1.1.2-) can still be searched for without a KO number (yes according to Frank bc the sequence exists but complicated by catabolism/anabolism))]

@@

[Anitra: OK to comment on this but needs to be 3-4 sentences] Taurine was also among the compounds highly enriched in the anomalous surface sample from Station 11 (MS11C215m_A), likely along with its trimethylated form taurine betaine. While putative in our analysis, trimethyltaurine has been previously reported in corals and red algae, where it probably has a destabilizing influence on proteins [@Impellizzeri1975; @Hill2017; @Bruzdziak2018]. We have a few hypotheses for the striking pattern in these compounds and the other 13 metabolites detailed above in Results. Most simply, the signal could be contamination during the sampling process, with some unknown material dominated by these compounds added to the filter during collection or processing. Alternatively, the signal could represent a real biological difference, perhaps indicating a pathway or organism highly abundant in this sample alone. As noted above, the MS11C215m_A sample was collected during a *Trichodesmium* bloom, an environment characterized by large heterogeneity even between replicate samples [@Rouco2014; @Rouco2016]. 10 liters of seawater were filtered for each sample but it is conceivable that a tuft or puff of *Trichodesmium* was captured in only this one as previously noted by @Rouco2014. The presence of a predictable trend across multiple chemically similar metabolites is the strongest bit of evidence in favor of the *Trichodesium* hypothesis, but the fact that these metabolites do not appear robustly in the bloom samples and that homoserine betaine did not appear to be especially abundant challenges it. One final note is that homoserine betaine was only documented as the sole major compatible solute in the Clade III *T. erythraeum* strain IMS101, while the bloom was more likely due to Clade I species given their relative abundance in the NPSG [@Rouco2016]. [Email Sonya about the presence of compatible solute genes in Clade I tricho? Especially taurine/TMT? Ask about IFCB samples (they're on the dashboard http://ifcbdb.soest.hawaii.edu/timeline?dataset=MESOSCOPE&bin=D20170708T043509_IFCB000 I just don't know what to ask)]

@@

A final pattern worth noting is the apparent upregulation of metabolites in the urea-ornithine cycle (OUC) among the samples taken during the Falkor cruise. Aspartate, citrulline, ornithine, and malic acid were all found to be more abundant in the Falkor samples relative to MESOSCOPE. The OUC is a staple of metazoan metabolism and is present in stramenopiles as well where it enables rapid response to increases in nitrogen supply [@Allen2011; @Bender et al?]. The enrichment of these metabolites in the Falkor cruise may indicate that there was still rapid uptake and turnover of nitrogen in the 2018 eddy pair while the 2017 eddies were more stable, despite both eddies being in a similar state of stable anticyclone and weakening cyclone [@Dugenne2023].
